Historical trends in the contribution of six major causes of death to the black-white

### 2017-04-26 Manuscript/Appendix-agestd-mort-rates.Rmd
better appendix

---
title: "Appendix: Trends in age-adjusted mortality rates for blacks and whites by cause and sex, 1969-2013, United States"
output: html_document
---

Corinne A Riddell, PhD

Department of Epidemiology, Biostatistics and Occupational Health, McGill University, Montreal, Canada.

@@

  a <- ggplot(subset.data, aes(x = year, y = rate.per.100k_mean)) + 
              geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, fill = Race), col = NA, alpha = 0.5) +
              geom_line(aes(col = Race)) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab(" ") + 
              ylab("Age-standardized mortality rate (per 100,000)") + 
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              ggtitle(paste0("Age-adjusted mortality (per 100,000) in ", sex1, " for ", COD1)) +
              theme_classic(base_size = 10) +
              theme(axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank())

  b <- ggplot(subset.data2, aes(x = year, y = rate.difference_mean)) +
           # geom_hline(aes(yintercept = -300), lwd = 0.5, col = "grey") +
           # geom_hline(aes(yintercept = -200), lwd = 0.5, col = "grey") +
           # geom_hline(aes(yintercept = -100), lwd = 0.5, col = "grey") +
           # geom_hline(aes(yintercept = 0), lwd = 0.5, col = "grey") +
           # geom_hline(aes(yintercept = 200), lwd = 0.5, col = "grey") +
           # geom_hline(aes(yintercept = 100), lwd = 0.5, col = "grey") +
              geom_line() +
              geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), col = NA, alpha = 0.5) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab(" ") + 
              ylab("Excess mortality among blacks (per 100,000)") +
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
              theme_classic(base_size = 10) +
              theme(panel.grid.minor = element_blank(),
                    axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank())

    return(list(plot1 = a, plot2 = b))
}
```

@@

**eFigure 1**

```{r print-em, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cardiovascular.Male.mort.plot[[1]]) %>% 
  layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F)

ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% 
  layout(xaxis = list(title = "Year (1969-2013)"))

#manipulating hover info:
#p <- ggplotly(Cardiovascular.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
#p$x$data[[1]]$hoverinfo = "none" #if want no hover info for a particular trace
#p$x$data[[1]]$text #otherwise, modified what is stored here to display how you want it to look. 
```

**eFigure 2**

```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cardiovascular.Female.mort.plot[[1]]) %>% 
  layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F)

ggplotly(Cardiovascular.Female.mort.plot[[2]]) %>% 
  layout(xaxis = list(title = "Year (1969-2013)"))
```

**eFigure 3**

```{r print-em-3, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cancers.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(Cancers.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
```

**eFigure 4**

```{r print-em-4, fig.width = 9, fig.height = 9, echo = F}
ggplotly(Cancers.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(Cancers.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
```

```{r print-3, fig.width = 9, fig.height=9, echo = F}
ggplotly(Communicable.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(Communicable.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))

ggplotly(Communicable.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(Communicable.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))

ggplotly(`Non-communicable.Male.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(`Non-communicable.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))

ggplotly(`Non-communicable.Female.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(`Non-communicable.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))

ggplotly(Injuries.Male.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(Injuries.Male.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))

ggplotly(Injuries.Female.mort.plot[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(Injuries.Female.mort.plot[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))

ggplotly(`All other causes.Male.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(`All other causes.Male.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))

ggplotly(`All other causes.Female.mort.plot`[[1]]) %>% layout(xaxis = list(title = "Year (1969-2013)"), showlegend = F) 
ggplotly(`All other causes.Female.mort.plot`[[2]]) %>% layout(xaxis = list(title = "Year (1969-2013)"))
```

>>>>

---
title: "Supplementary Appendix: Trends in age-standardized mortality rates for blacks and whites by cause and sex, 1969-2013, United States"
output: html_document
---

This is the supplementary appendix for the paper, "Historical trends in the contribution of six major causes of death to the black-white life expectancy gap by US state."

Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD

@@

  a <- ggplot(subset.data, aes(x = year, y = rate.per.100k_mean)) + 
              geom_ribbon(aes(ymin = rate.per.100k_lcl, ymax = rate.per.100k_ucl, group = Race), fill = "grey", col = NA, alpha = 0.5) +
              geom_line(aes(col = Race)) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab("Year (1969-2013)") + 
              ylab("Age-standardized mortality rate (per 100,000)") + 
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              theme_classic(base_size = 10) +
              theme(axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank())

  b <- ggplot(subset.data2, aes(x = year, y = rate.difference_mean)) +
              geom_line() +
              geom_ribbon(aes(ymin = rate.difference_LCL, ymax = rate.difference_UCL), col = NA, alpha = 0.5) +
              facet_wrap( ~ stabbrs.map.order, ncol = 11, drop = F) +
              xlab("Year (1969-2013)") + 
              ylab("Excess mortality among blacks (per 100,000)") +
              geom_hline(aes(yintercept = 0)) + geom_vline(aes(xintercept = 1969)) +  
              #ggtitle(paste0("Excess mortality among blacks (per 100,000) in ", sex1, " for ", COD1)) +
              theme_classic(base_size = 10) +
              theme(panel.grid.minor = element_blank(),
                    axis.text.x = element_blank(),
                    strip.background=element_blank(),
                    axis.line=element_blank(),
                    axis.ticks=element_blank())

  c <- ggplotly(a)
  d <- ggplotly(b)

  for(i in 1:length(c$x$data)){
    if (c$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
      c$x$data[[i]]$hoverinfo <- "none"
    }

    c$x$data[[i]]$text <- gsub("rate.per.100k_mean", "Mean mortality rate", c$x$data[[i]]$text)
    c$x$data[[i]]$text <- gsub("rate.per.100k_lcl", "Lower credible limit", c$x$data[[i]]$text)
    c$x$data[[i]]$text <- gsub("rate.per.100k_ucl", "Upper credible limit", c$x$data[[i]]$text)
  }

  for(i in 1:length(d$x$data)){
    if (d$x$data[[i]]$line$color == "rgba(0,0,0,1)") {
      d$x$data[[i]]$hoverinfo <- "none"
    }

    d$x$data[[i]]$text <- gsub("rate.difference_mean", "Mean excess mortality", d$x$data[[i]]$text)
    d$x$data[[i]]$text <- gsub("rate.difference_LCL", "Lower credible limit", d$x$data[[i]]$text)
    d$x$data[[i]]$text <- gsub("rate.difference_UCL", "Upper credible limit", d$x$data[[i]]$text)
  }

    return(list(plot1 = a, plot2 = b, plot3 = c, plot4 = d))
}
```

@@

**eFigure 1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**

```{r print-em, fig.width = 9, fig.height = 9, echo = F}
Cardiovascular.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
Cardiovascular.Male.mort.plot[[4]]
```

**eFigure 2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**

```{r print-em-2, fig.width = 9, fig.height = 9, echo = F}
Cardiovascular.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
Cardiovascular.Female.mort.plot[[4]]
```

**eFigure 3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**

```{r print-em-3, fig.width = 9, fig.height = 9, echo = F}
Cancers.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
Cancers.Male.mort.plot[[4]]
```

**eFigure 4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**

```{r print-em-4, fig.width = 9, fig.height = 9, echo = F}
Cancers.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
Cancers.Female.mort.plot[[4]]
```

**eFigure 5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 9, echo = F}
`Non-communicable.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100))
`Non-communicable.Male.mort.plot`[[4]]
```

**eFigure 6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 9, echo = F}
`Non-communicable.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100))
`Non-communicable.Male.mort.plot`[[4]]
```

**eFigure 7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 9, echo = F}
Injuries.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
Injuries.Male.mort.plot[[4]]
```

**eFigure 8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 9, echo = F}
Injuries.Female.mort.plot[[3]] %>% layout(margin = list(b = 100)) 
Injuries.Female.mort.plot[[4]]
```

**eFigure 9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**

```{r print-3, fig.width = 9, fig.height=9, echo = F}
Communicable.Male.mort.plot[[3]] %>% layout(margin = list(b = 100))
Communicable.Male.mort.plot[[4]]
```

**eFigure 10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**

```{r print-4, fig.width = 9, fig.height=9, echo = F}
Communicable.Female.mort.plot[[3]] %>% layout(margin = list(b = 100))
Communicable.Female.mort.plot[[4]]
```

**eFigure 11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 9, echo = F}
`All other causes.Male.mort.plot`[[3]] %>% layout(margin = list(b = 100))
`All other causes.Male.mort.plot`[[4]]
```

**eFigure 12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**

```{r, fig.width = 9, fig.height = 9, echo = F}
`All other causes.Female.mort.plot`[[3]] %>% layout(margin = list(b = 100)) 
`All other causes.Female.mort.plot`[[4]]
```

### 2017-04-26 Manuscript/Manuscript_COD_Trends.Rmd
better appendix

Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure X**). Recently, there has been a slight decrease in mortality rates, especially among black males. These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure Y**). As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with previous causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.

@@

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure X**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure X**). Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases. Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available. This spike was greater among males, especially blacks, and within states with larger urban centers. Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks. White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

>>>>

Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure 5**). Recently, there has been a slight decrease in mortality rates, especially among black males. These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure 6**). As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with previous causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.

@@

Injury is one of the least common causes of death. Over time, injury-related mortality has decreased precipitously among blacks, although an [insurgence][better word?] of crime resulted in heightened mortality from injuries among black males during the 1990s (**eFigure 7**). For whites, most states exhibit an initial period of gradual decreases in injury mortality, followed by a period of gradual increases, especially among white females (**eFigure 8**). Thus, injury mortality rates among blacks are the same as, or lower than those among whites in many states, with white females demonstrating sharper increases. Thus, the contribution of injuries to the mortality gap has generally decreased over time, and has become negative in those states with lower injury-related mortality among blacks than whites (**Figures 9 and 10**). Among males, DC, Illinois, Michigan and Missouri exhibit the largest recent injury-related contributions. Among females, many southern states exhibit negative contributions (indicating lower injury-related mortality among black women), while DC, Illinois, and Michigan having higher contributions than the general trend.

@@

Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available (**eFigures 9 and 10**). This spike was greater among males, especially blacks, and within states with larger urban centers. Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks (**eFigures 11 and 12**). White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.
